---
title: "01- Descriptive Analysis"
author: "EDA 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Set working directory either using point-clink method or programming setwd()
```{r}
getwd()
#("C:/Users/redbe/OneDrive/Documents/Data")
#setwd("~/Data")
```

#import data
```{r}
pd<-read.csv("Parkinson.csv")
#pd<-read.csv(file.choose())
```

#Description of data
#Matrix column entries (attributes):
# name - ASCII subject name and recording number
#MDVP:Fo(Hz) - Average vocal fundamental frequency
#MDVP:Fhi(Hz) - Maximum vocal fundamental frequency
#MDVP:Flo(Hz) - Minimum vocal fundamental frequency
#MDVP:Jitter(%),MDVP:Jitter(Abs),MDVP:RAP,MDVP:PPQ,Jitter:DDP - Several 
#measures of variation in fundamental frequency
#MDVP:Shimmer,MDVP:Shimmer(dB),Shimmer:APQ3,Shimmer:APQ5,MDVP:APQ,Shimmer:DDA - Several measures of variation in amplitude
#NHR,HNR - Two measures of ratio of noise to tonal components in the voice
#status - Health status of the subject (one) - Parkinson's, (zero) - healthy
#RPDE,D2 - Two nonlinear dynamical complexity measures
#DFA - Signal fractal scaling exponent
#spread1,spread2,PPE - Three nonlinear measures of fundamental frequency variation 

```{r}
View(pd)
head(pd) #first 6 rows
tail(pd) #last 6 rows
```

```{r}
#reviewing label names in the dataset
names(pd)
```
```{r}
#structure
str(pd)

```

```{r}
unique(pd$status)
```

```{r}
pd$status <- as.factor(pd$status) #now a categorical variable
str(pd) #confirm structure of pd 
```

#Creating subset of data using subset function
```{r}
counts <- subset(pd,RPDE>=.5)
```

#package pillar
```{r}
library("pillar")
glimpse(pd)
```

```{r}
#is.na(pd)
#counting is missing values
sum(is.na(pd))
sum(is.na(pd$MDVP.Fo.Hz.))
```

#summary statistics of Parkinson dataset or by specific column
#measures of central tendency
```{r}
summary(pd)
summary(pd$status)
```

#summary by group
```{r}
by(pd$MDVP.Fo.Hz., pd$status, summary)
```

#skewness and kurtosis
#measures of dispersion
```{r}
library("e1071")
skewness(pd$spread1)
plot(density(pd$spread1))

kurtosis(pd$MDVP.Flo.Hz.)
plot(density(pd$MDVP.Flo.Hz.))
```

#frequency table and proportions
```{r}
table(pd$status)

#proportions of the table
round(proportions(table(pd$status)),3)
```

#Go to the knit button, choose "Knit to HMTL"
#The HTML document will save in the working directory

